Blame and Coercion: Together Again for the Second Time

### 2019-10-08 paper.tex

\newtheorem{theorem}{Theorem}[]
\newtheorem{proposition}{Proposition}[]
\newtheorem{corollary}{Corollary}[]
\newcommand{\figref}[1]{Fig.~\ref{#1}}
\newcommand{\stxrule}[3]{#1 & ::= & #3 & \text{#2}\\}
\newcommand{\stxrulecont}[1]{& | & #1 & \\}
\newcommand{\funrule}[3]{#1 & = & #2 & #3\\}
\newcommand{\comprule}[4]{#1 & \fatsemi^l & #2 & = & #3 & #4 \\}
\newcommand{\plus}[0]{+}
\newcommand{\judgetype}[3]{#1 \vdash #2 : #3}
\newcommand{\lazyUD}{$\mathtt{Lazy UD}$}
\newcommand{\lazyD}{$\mathtt{Lazy D}$}
\newcommand{\eager}{$\mathtt{Eager}$}

@@

\section{Blame Calculus} \label{sec:blame-calculus}

@@

One might be tempted to conclude that inconsistency is the root of all runtime 
type errors (blames). In the setting of gradual typing, however, runtime 
type-checking is usually gradual as well. 
$ S \smile T $ reads ``$ S $ and $ T $ are shallowly-consistent''. Two types 
are shallowly consistent if one of them is $ \star $, or they have the same 
top-most type constructor. Shallow-inconsistency is the root of all blames -- 
casting a value to a shallowly inconsistent type leads to a blame immediately. 
Shallow-consistency is also reflexive, symmetric, and not transitive.

@@

We describe the dynamic semantics of \lazyD{} Blame Calculus with the CEK 
machine (\citet{felleisen1986control}). We choose CEK machine because to 
achieve space efficiency, we need to change continuations. 
Thus, abstract machines that include continuation in states is quite convenient 
for us. And CEK is one of the simplest machine of this kind.

@@

\section{Coercion Calculus}

@@

\begin{figure}
	\[ 
	\begin{array}{lclr}
	\stxrule{c}{hyper-coercions}{
		id* \mid{}
		( h , ( b , t ) )
	}
	\stxrule{h}{heads}{
		\epsilon \mid{}
		?^l
	}
	\stxrule{m}{middles}{
		U \mid{}
		c_1 \rightarrow c_2 \mid{}
		c_1 \times c_2 \mid{}
		c_1 \plus c_2
	}
	\stxrule{t}{tails}{
		\epsilon \mid{}
		! \mid{}
		\bot^l
	}
	\end{array}
	\]

	\fbox{$ c \fatsemi^l c = c $}
	\[ 
	\begin{array}{rclclr}
	\comprule{id*}{id*}{
		id*
	}{}
	\comprule{id*}{( ?^{l'} , ( b , t ) )}{
		( ?^{l'} , ( b , t ) )
	}{}
	\comprule{id*}{( ! , ( b , t ) )}{
		(?^{l} , ( b , t))
	}{}
	\comprule{(h , (b ,\bot^{l'}))}{c}{
		(h , (b ,\bot^{l'}))
	}{}
	\comprule{(h , (b_1 , t_1))}{id*}{
		(h , (b_1 , !))
	}{t_1 \neq \bot^{l'}}
	\comprule{(h , (b_1 , t_1))}{(\epsilon , (b_2 , t_2))}{
		(h , b_1 \fatsemi^{l} (b_2 , t_2))
	}{t_1 \neq \bot^{l'}}
	\comprule{(h , (b_1 , t_1))}{(?^{l'} , (b_2 , t_2))}{
		(h , b_1 \fatsemi^{l'} (b_2 , t_2))
	}{t_1 \neq \bot^{l'}}
	\end{array}
	\]

	\fbox{$ b \fatsemi^l (b,t) = (b,t) $}
	\[ 
	\begin{array}{rclclr}
	\comprule{U}{(U,t)}{
		(U,t)
	}{}
	\comprule{c_1 \rightarrow c_2}{(c_3 \rightarrow c_4,t)}{
		(c_3 \fatsemi^{l} c_1 \rightarrow c_2 \fatsemi^l c_4, t)
	}{}
	\comprule{c_1 \times c_2}{(c_3 \times c_4,t)}{
		(c_1 \fatsemi^{l} c_3 \times c_2 \fatsemi^l c_4, t)
	}{}
	\comprule{c_1 \plus c_2}{(c_3 \plus c_4,t)}{
		(c_1 \fatsemi^{l} c_3 \plus c_2 \fatsemi^l c_4, t)
	}{}
	\comprule{b_1}{(b_2,t)}{
		(b_1,\bot^l)
	}{b_1 }
	\end{array}
	\]

	\fbox{$ seq(c,c) = c $}
	\[
	\begin{array}{rclr}
	\funrule{seq(c_1,c_2)}{
		c_1 \fatsemi^l c_2
	}{}
	\end{array}
	\]

	\fbox{$ id( T ) = c $}
	\[
	\begin{array}{rclr}
	\funrule{id(*)}{
		id*
	}{}
	\funrule{id(P)}{
		(\epsilon,(id(P),\epsilon))
	}{}
	\end{array}
	\]

	\fbox{$ id( P ) = b $}
	\[
	\begin{array}{rclr}
	\funrule{id(U)}{U}{}
	\funrule{id(T_1 \rightarrow T_2)}{
		id(T_1) \rightarrow id(T_2)
	}{}
	\funrule{id(T_1 \times T_2)}{
		id(T_1) \times id(T_2)
	}{}
	\funrule{id(T_1 \plus T_2)}{
		id(T_1) \plus id(T_2)
	}{}
	\end{array}
	\]

>>>>

\newtheorem{theorem}{Theorem}[]
\newtheorem{proposition}{Proposition}[]
\newtheorem{corollary}{Corollary}[]
\newcommand{\figref}[1]{Fig.~\ref{#1}}
\newcommand{\stxrule}[3]{#1 & ::= & #3 & \text{#2}\\}
\newcommand{\stxrulecont}[1]{& | & #1 & \\}
\newcommand{\funrule}[3]{#1 & = & #2 & #3\\}
\newcommand{\comprule}[4]{#1 & \fatsemi^\ell & #2 & = & #3 & #4 \\}
\newcommand{\plus}[0]{+}
\newcommand{\judgetype}[3]{#1 \vdash #2 : #3}
\newcommand{\lazyUD}{$\mathtt{Lazy UD}$}
\newcommand{\lazyD}{$\mathtt{Lazy D}$}
\newcommand{\eager}{$\mathtt{Eager}$}
\newcommand{\hyperCoercionI}[0]{id\star}
\newcommand{\hyperCoercionC}[3]{#1 \overset{#2}{\curvearrowright} #3}

@@

\section{Background: Blame Calculus} \label{sec:blame-calculus}

@@

One might be tempted to conclude that inconsistency is the root of all runtime 
type errors (blames). In the setting of gradual typing, however, runtime 
type-checking is usually gradual as well. 
$ S \smile T $ reads ``$ S $ and $ T $ are shallowly-consistent''. Two types 
are shallowly consistent if one of them is $ \star $, or they have the same 
top-most type constructor. Shallow-inconsistency is the root of all blames in 
all lazy strategies -- casting a value to a shallowly inconsistent type leads 
to a blame immediately. Shallow-consistency is also reflexive, symmetric, and 
not transitive.

@@

All existing proofs describe the semantics of \lazyD{} Blame Calculus in the 
style of reduction semantics. 
We instead do it with the CEK machine (\citet{felleisen1986control}). 
We choose CEK machine because to achieve space efficiency, we need to change 
continuations. Thus, abstract machines that include continuation in states is 
quite convenient for us. And CEK is one of the simplest machine of this kind.

@@

\section{Background: Coercion Calculus}

@@

\begin{figure}
	\[ 
	\begin{array}{lclr}
	\stxrule{c}{hyper-coercions}{
		\hyperCoercionI \mid{}
		\hyperCoercionC{h}{m}{t}
	}
	\stxrule{h}{heads}{
		\epsilon \mid{}
		?^l
	}
	\stxrule{m}{middles}{
		U \mid{}
		c_1 \rightarrow c_2 \mid{}
		c_1 \times c_2 \mid{}
		c_1 \plus c_2
	}
	\stxrule{t}{tails}{
		\epsilon \mid{}
		! \mid{}
		\bot^l
	}
	\stxrule{\ell}{$ \mathtt{Maybe} \; l $}{
		\epsilon \mid{}
		l
	}
	\end{array}
	\]

	\fbox{$ c \fatsemi^\ell c = c $}
	\[ 
	\begin{array}{rclclr}

	\comprule{
		\hyperCoercionI
	}{
		\hyperCoercionI
	}{
		\hyperCoercionI
	}{}

	\comprule{
		\hyperCoercionI
	}{
		\hyperCoercionC{?^{l'}}{b}{t}
	}{
		\hyperCoercionC{?^{l'}}{b}{t}
	}{}

	\comprule{
		\hyperCoercionI
	}{
		\hyperCoercionC{!}{b}{t}
	}{
		\hyperCoercionC{?^{l}}{b}{t}
	}{\ell = l}

	\comprule{
		\hyperCoercionC{h}{b}{\bot^{l'}}
	}{
		c
	}{
		\hyperCoercionC{h}{b}{\bot^{l'}}
	}{}

	\comprule{
		\hyperCoercionC{h}{b_1}{t_1}
	}{
		\hyperCoercionI
	}{
		\hyperCoercionC{h}{b_1}{!}
	}{
		t_1 \neq \bot^{\centerdot}
	}

	\comprule{
		\hyperCoercionC{h}{b_1}{t_1}
	}{
		\hyperCoercionC{\epsilon}{b_2}{t_2}
	}{
		\hyperCoercionC{h}{b'}{t'}
	}{
		t_1 \neq \bot^{\centerdot} \; \text{and} \;
		b_1 \fatsemi^{\ell} (b_2, t) = (b', t')
	}

	\comprule{
		\hyperCoercionC{h}{b_1}{t_1}
	}{
		\hyperCoercionC{?^{l'}}{b_2}{t_2}
	}{
		\hyperCoercionC{h}{b'}{t'}
	}{
		t_1 \neq \bot^{\centerdot} \; \text{and} \;
		b_1 \fatsemi^{l'} (b_2, t) = (b', t')
	}
	\end{array}
	\]

	\fbox{$ m \fatsemi^\ell (m,t) = (m,t) $}
	\[ 
	\begin{array}{rclclr}
	\comprule{U}{(U,t)}{
		(U,t)
	}{}
	\comprule{c_1 \rightarrow c_2}{(c_3 \rightarrow c_4,t)}{
		(c_3 \fatsemi^{l} c_1 \rightarrow c_2 \fatsemi^\ell c_4, t)
	}{}
	\comprule{c_1 \times c_2}{(c_3 \times c_4,t)}{
		(c_1 \fatsemi^{l} c_3 \times c_2 \fatsemi^\ell c_4, t)
	}{}
	\comprule{c_1 \plus c_2}{(c_3 \plus c_4,t)}{
		(c_1 \fatsemi^{l} c_3 \plus c_2 \fatsemi^\ell c_4, t)
	}{}
	\comprule{b_1}{(b_2,t)}{
		(b_1,\bot^l)
	}{
		\ell = l \; \text{and} \;
		\neg b_1 \smile b_2 
	}
	\end{array}
	\]

	\fbox{$ seq(c,c) = c $}
	\[
	\begin{array}{rclr}
	\funrule{seq(c_1,c_2)}{
		c_1 \fatsemi^\epsilon c_2
	}{cod(c_1) = dom(c_2)}
	\end{array}
	\]

	\fbox{$ id( P ) = m $}
	\[
	\begin{array}{rclr}
	\funrule{id(U)}{U}{}
	\funrule{id(T_1 \rightarrow T_2)}{
		id(T_1) \rightarrow id(T_2)
	}{}
	\funrule{id(T_1 \times T_2)}{
		id(T_1) \times id(T_2)
	}{}
	\funrule{id(T_1 \plus T_2)}{
		id(T_1) \plus id(T_2)
	}{}
	\end{array}
	\]

	\fbox{$ id( T ) = c $}
	\[
	\begin{array}{rclr}
	\funrule{id(\star)}{
		\hyperCoercionI
	}{}
	\funrule{id(P)}{
		\hyperCoercionC{\epsilon}{id(P)}{\epsilon}
	}{}
	\end{array}
	\]
